An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-11-01 scagnostics.Rmd

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the scagnostics on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are:

- **pipe**: $n$ points where all $p$ directions are independently drawn from a uniform distribution between $[-1,1]$, but enforcing $x_p^2 + x_{p-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- **sine**:  $n$ points where $p-1$ directions are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_p = \sin(x_{p-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"
- **spiral**:  $n$ points where $p-2$ directions are independently drawn from a normal distribution, and direction $p-1$ and $p$ are sampled on an Archimedean spiral, i.e. $r = a + b \theta$, with $a=b=0.1$ and we sample angles $\theta$ from a normal distribution with mean 0 and variance $2\pi$, giving a spiral with higher densities at lower radii. We take the absolute value of $\theta$ to fix the direction of the spiral shape when sampling the points.

>>>>

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the scagnostics on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $p$ is used for the number of points, $n$ is the number of dimensions, and $d=2$ is the projection dimension.

- **pipe**: $p$ points where $n-1$ points are independently drawn from a uniform distribution between $[-1,1]$, and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- **sine**:  $p$ points where $n-1$ points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"
- **spiral**:  $n$ points where $p-2$ directions are independently drawn from a normal distribution, and direction $p-1$ and $p$ are sampled on an Archimedean spiral, i.e. $r = a + b \theta$, with $a=b=0.1$ and we sample angles $\theta$ from a normal distribution with mean 0 and variance $2\pi$, giving a spiral with higher densities at lower radii. We take the absolute value of $\theta$ to fix the direction of the spiral shape when sampling the points.
